---
title: "Hands-on Exercise 1: A Layered Grammar of Graphics"
author: "Vanessa Heng"
date: "January 9, 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  #freeze: true
  warning: false
  message: false
editor: visual
---

# 1 Getting Started

## 1.1 Installing and launching R packages

The code below uses p_load() of the pacman package to check if tidyverse packages are installed in the laptop. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## 1.2 Importing the data

Let's use read_csv function of [readr](https://readr.tidyverse.org/) package to load the data from *Exam_data.csv* into R.

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# 2 Exploratory Data Analysis

Let's understand more of the data given.

```{r}
summary(exam_data)
```

```{r}
head(exam_data, 5)
```

::: callout-tip
## Observations

1.  This data is about students' examination scores for their 3 subjects (English, Maths, and Science) from various classes in Primary 3 from a school.

2.  There are 4 categorical variables: ID, CLASS, GENDER, and RACE, and 3 continuous variables: ENGLISH, MATHS, and SCIENCE.

3.  Out of the 3 subjects, Maths has the higher mean and median score.
:::

## 2.1 Examination scores distribution

Let's create boxplots for the performance of 3 subjects and arrange them in a row for easy comparison.

```{r}
library("ggpubr")
boxplt1 <- ggplot(data = exam_data,   
  aes(y = MATHS)) +   
  geom_boxplot() 
boxplt2 <- ggplot(data = exam_data,   
  aes(y = ENGLISH)) +   
  geom_boxplot()
boxplt3 <- ggplot(data = exam_data,   
  aes(y = SCIENCE)) +   
  geom_boxplot()
ggarrange(boxplt1, boxplt2, boxplt3, ncol = 3, nrow = 1)
```

::: callout-tip
## Observation

Students did better in Maths as compared to English and Science. There are some outliers in Maths and English boxplots where students did very poorly.
:::

Let's dive a little deeper to see how each class performs.

```{r}
ggplot(data = exam_data,   
  aes(x = CLASS, y = MATHS)) +   
  geom_boxplot()+
  stat_summary(geom = "point",       
               fun = "mean",         
               colour = "red",        
               size = 3) 
ggplot(data = exam_data,   
  aes(x = CLASS, y = ENGLISH)) +   
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour = "red",        
               size = 3) 
ggplot(data = exam_data,   
  aes(x = CLASS, y = SCIENCE)) +   
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour = "red",        
               size = 3) 
```

::: callout-tip
## Observations

1.  There are 9 classes in this dataset.

2.  Students from Class 3I have poorer performance in all 3 subjects (many failed 3 subjects), whereas students from Class 3A have better performance in all 3 subjects.
:::

## 2.2 Race distribution

Let's take a look at the students' race distribution:

```{r}
ggplot(data = exam_data, 
       aes(x = fct_infreq(RACE))) +
  geom_bar() +
  labs(title="Distribution of race",
       x = "Race",
       y = "Number of students") + 
  ylim(0, 210) +
  geom_text(aes(label = after_stat(count)), 
            stat = "count", vjust = -0.5, colour = "brown")
```

::: callout-tip
## Observation

From the bar chart above, the majority of the students are Chinese, followed by Malay.
:::

What about the race distribution in each class?

```{r}
ggplot(data = exam_data) +
  geom_bar(mapping = aes(x =CLASS, fill = RACE))

```

::: callout-tip
## Observations

1.  Class 3A has more Chinese students whereas Class 3I has more Malay students.

2.  Class 3I has the fewest number of students.
:::

## 2.3 Gender distribution

Let's take a look at the students' gender distribution:

```{r}

ggplot(data = exam_data, 
       aes(x = fct_infreq(GENDER))) +
  geom_bar() +
  labs(title="Distribution of gender",
       x = "Gender",
       y = "Number of students") + 
  ylim(0, 210) +
  geom_text(aes(label = after_stat(count)), 
            stat = "count", vjust = -0.5, colour = "brown")
```

::: callout-tip
## Observation

There are more female students than male students.
:::

What about the distribution of gender in each class?

```{r}
ggplot(data = exam_data) +
  geom_bar(mapping = aes(x =CLASS, fill = GENDER))
```

::: callout-tip
## Observation

There are more female students in Class 3B and fewer female students in Class 3I.
:::

## 2.4 Correlation between performances of different subjects

Let's see if there is any correlation between the performance of Maths vs English, Maths vs Science, and English vs Science via scatter plots.

```{r}

ggplot(data=exam_data,
  aes(x = MATHS,y = ENGLISH)) +
  geom_point() + 
  geom_smooth(linewidth = 0.5) +
  # use the same axis range
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

ggplot(data=exam_data,
  aes(x = MATHS,y = SCIENCE)) +
  geom_point() + 
  geom_smooth(linewidth = 0.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))

ggplot(data=exam_data,
  aes(x = ENGLISH, y = SCIENCE)) +
  geom_point() + 
  geom_smooth(linewidth = 0.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

::: callout-tip
## Observations

All 3 scatter plots show some positive correlation between the performances of the subjects. However, we would need to calculate the correlation coefficients to determine how strong the correlations are.
:::
