---
title: "In-class Exercise 1: Now you see it!"
author: "Vanessa Heng"
date: "January 13, 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  #freeze: true
  warning: false
  message: false
editor: visual
---

## Loading R packages

In this In-class exercise, two R packages will be used. They are

-   [tidyverse](https://www.tidyverse.org/)

-   [haven](https://haven.tidyverse.org/)

The code chunk used is as follows:

```{r}
pacman::p_load(tidyverse, haven)
```

## Importing PISA data

The code chunk below uses [read_sas()](https://haven.tidyverse.org/reference/read_sas.html) of [haven](https://haven.tidyverse.org/reference/index.html) to import PISA data into the R environment.

```{r}
#| eval: false
stu_qqq <- read_sas("data/cy08msp_stu_qqq.sas7bdat")
```

Let's filter data by Country ID = SGP (Singapore).

```{r}
#| eval: false
stu_qqq_SG <- stu_qqq %>%
  filter(CNT == "SGP")
```

## Filtering and saving filtered data

Let's save the filtered data into a file (for future easy loading).

```{r}
#| eval: false
write_rds(stu_qqq_SG, "data/stu_qqq_SG.rds")
```

The following code is to read an RDS file (i.e. the saved filtered (by SG) file).

```{r}
stu_qqq_SG <- read_rds("data/stu_qqq_SG.rds")
```
